---
title: "notes"
author: "Vajira"
date: "2025-06-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include = FALSE}
library(dplyr)
library(ggplot2)
library(mapview)
library(leaflet)
library(leafpop)
library(sf)

sites <- read.csv('D:/src/flux-rootzone/data/fdk_site_info.csv') |>
  rename(
    'koeppen' = 'koeppen_code',
    'start' = 'year_start',
    'end' = 'year_end',
    'IGBP' = 'igbp_land_use'
  )
```

# Ecosystem-water interactions at flux sites

We use [FluxDataKit](https://doi.org/10.5281/zenodo.10885933) which gives a compilation of flux data, meterological data and remote sensing vegetation data (MODIS LAI and fAPAR) to explore ecosystem-water dynamics with a focus on green water storage capacities (Root zone water storage capacity). Data reference [Ukkola et al, (2022)](https://doi.org/10.5194/essd-14-449-2022).

## Data Coverage

```{r echo = FALSE}
leaflet(sites) |>
  addProviderTiles(providers$Esri.WorldImagery) |>
  addMarkers(~lon, ~lat, label = ~sitename, popup = popupTable(sites))
```

The dataset is generated compiling multiple sources of flux data.

Coverage varies across Depending on the source of the data the included land cover types vary. For example Oneflux is dominated by mostly cropland sites. Land cover labels are based on MODIS.LC classes are IGBP.[Reference for IGBP land cover types.](https://www.ceom.ou.edu/static/docs/IGBP.pdf)

```{r echo = FALSE}
# plot the number of sites by IGBP LC type as a bar chart
sites %>%
  group_by(IGBP) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = reorder(IGBP, -n), y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "IGBP Land Cover Type", y = "Number of Sites", title = "Number of Sites by IGBP Land Cover Type") +
  theme_minimal()
```

Coverage varies by climate
```{r echo = FALSE}
# plot the number of sites by Koeppen climate type as a bar chart
sites %>%
  group_by(koeppen) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = reorder(koeppen, -n), y = n)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "Koeppen Climate Type", y = "Number of Sites", title = "Number of Sites by Koeppen Climate Type") +
  theme_minimal()
```
## Root zone water storage capacity
Root zone water storage capacity at the sites referred from [Stocker et al, (2023)](https://www.nature.com/articles/s41561-023-01125-2) included as site metadata. 

```{r echo = FALSE}
# plot the distribution of root zone water storage capacity in the sites as a histogram
sites %>%
  ggplot(aes(x = whc)) +
  geom_histogram(binwidth = 25, fill = "blue", color = "black") +
  labs(x = "Root Zone Water Storage Capacity (mm)", y = "Frequency", title = "Distribution of Root Zone Water Storage Capacity") +
  theme_minimal()
```

```{r echo = FALSE}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)

# Load site data from FluxDataKit
# Assuming fdk_site_fullyearsequence is already loaded in your environment
# If not, you would load it with:
# load("path/to/fdk_site_fullyearsequence.rda")

# If we need to recreate it from the CSV (as shown in the repo):
site_data <- read_csv("D:/src/flux-rootzone/data/fdk_site_info.csv")

# Alternatively, if you have the package installed:
# data("fdk_site_fullyearsequence", package = "FluxDataKit")
# site_data <- fdk_site_fullyearsequence

# Create Budyko curve function
budyko_curve <- function(ai) {
  # Original Budyko (1974) formulation
  sqrt(ai * tanh(1/ai) * (1 - exp(-ai)))
}

# Prepare data for plotting
plot_data <- site_data %>%
  filter(!is.na(p_over_pet)) %>%  # Remove sites with missing P/PET
  mutate(budyko_et_ratio = budyko_curve(1/p_over_pet))  # Calculate Budyko ET ratio

# Create Budyko plot
budyko_plot <- ggplot() +
  # Budyko curve
  stat_function(fun = budyko_curve, 
                aes(color = "Budyko Curve"), 
                size = 1.2,
                xlim = c(0.1, 10)) +
  
  # 1:1 line (energy limit)
  geom_abline(slope = 1, intercept = 0, 
              linetype = "dashed", 
              color = "gray50") +
  
  # Horizontal line at 1 (water limit)
  geom_hline(yintercept = 1, 
             linetype = "dashed", 
             color = "gray50") +
  
  # Site points
  geom_point(data = plot_data,
             aes(x = 1/p_over_pet, 
                 y = 1,  # Assuming ET/P = 1 for simplicity - adjust if you have actual ET data
                 fill = igbp_land_use),  # Color by IGBP class
             size = 3, 
             shape = 21,
             alpha = 0.8) +
  
  # Site labels (for a subset to avoid overcrowding)
  geom_text_repel(data = plot_data %>% 
                    filter(p_over_pet < 0.5 | p_over_pet > 2),  # Label extreme sites
                  aes(x = 1/p_over_pet, 
                      y = 1,
                      label = sitename),
                  size = 3,
                  box.padding = 0.5,
                  max.overlaps = Inf) +
  
  # Aesthetics
  scale_x_log10(limits = c(0.1, 10),
                breaks = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
                labels = c("0.1", "0.2", "0.5", "1", "2", "5", "10")) +
  scale_y_continuous(limits = c(0, 1.2),
                     breaks = seq(0, 1.2, by = 0.2)) +
  scale_color_manual(name = "",
                     values = c("Budyko Curve" = "black")) +
  scale_fill_discrete(name = "IGBP Class") +
  
  # Labels
  labs(x = "Aridity Index (PET/P)",
       y = "Evaporative Index (ET/P)",
       title = "FluxDataKit Sites on Budyko Curve",
       subtitle = "Showing relationship between aridity and evaporation") +
  
  # Theme
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())

# Print the plot
print(budyko_plot)

# Save the plot
#ggsave("budyko_plot.png", budyko_plot, width = 10, height = 8, dpi = 300)

```
